---
title: "ObsTrans"
author: "Warren Coons"
date: "2024-02-17"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup - Packages and Working Directory

```{r}
## Function to load packages 
install_load <- function(packages){
  for (p in packages) {
    if (p %in% rownames(installed.packages())) {
      library(p, character.only=TRUE)
    } else {
      install.packages(p)
      library(p, character.only = TRUE)
    }
  }
}

######## INSTALL PACKAGES
package_list<-c("dplyr", "data.table", "haven", "stats")
install_load(package_list)
```

# General Exclusions (Applied to both NHANES and BRFSS)

- Those less than 20 yrs old
- Those with missing data in any covariates

## NHANES 

# Overview

Variables:

Exposure: LBXBPB = Lead level (lead)
Outcome: hyper = Hypertension (BP)
Covariates:
- Age: RIDAGEYR (demo) 
- Sex: RIAGENDR (demo)
- Race: RIDRETH1 (demo)
- Educational Attainment: DMDEDUC2 (demo)
- BMI: BMXBMI (BMI)
- Smoking behavior: SMQ020 (smoke) and SMQ40 (smoke)
- Alcohol use: ALQ130 (ALC)

List of possible EMM on the RD scale found in BRFSS:
- Stress: HSQ480 (maybe use depression screener) (stress)
- Sleep duration: SLD010H (sleep)

List of possible EMM on the RD scale NOT found in BRFSS:
- Diets heavy in salt (Given by sodium levels on day 1): XX NOT FOUND IN BRFSS (DR1TSODI)
- Inadequate consumption of fruits and vegetables: XX NOT FOUND IN BRFSS
- Body mass: XX CONFOUNDER
- Physical Activity: XX QUESTION DOESN'T HAVE AN ANALOG FOR BRFSS
- Trouble sleeping: XX NOT FOUND IN BRFSS
- Sleep disorders: XX NOT FOUND IN BRFSS

Codebooks:
- Lead: https://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/PBCD_G.htm
- BP: https://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/BPX_G.htm
- Demo: https://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/DEMO_G.htm
- BMI: https://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/BMX_G.htm
- Smoke: https://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/SMQ_G.htm
- ALC: https://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/ALQ_G.htm
- Stress: https://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/HSQ_G.htm#HSQ480
- Sleep: https://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/SLQ_G.htm

Risk Factors of Hypertension - Codebooks:
- (Stress) https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1940093/
- (Sleep) https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8345503/ (specifically sleeping duration, trouble sleeping, and sleep disorders)
- (Many others) https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10181897/

Study Inspiration:
- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9081331/#:~:text=In%20some%20studies%2C%20blood%20lead,demonstrated%20(19%2C%2020). (Several confounders and exclusion criteria come from this study)

# NHANES Setup - Importing Data

```{r}
# Exposure

lead <- read_xpt("../Data/Source/PB12.XPT")

# Outcome (to be converted to hypertension)
BP <- read_xpt("../Data/Source/BP12.XPT")

# Confounders
demo <- read_xpt("../Data/Source/DEMO12.XPT")
BMI <- read_xpt("../Data/Source/BMI12.XPT")
smoke <- read_xpt("../Data/Source/SMOKE12.XPT")
ALC <- read_xpt("../Data/Source/ALQ12.XPT")

# Possible EMM
stress <- read_xpt("../Data/Source/STRESS12.XPT")
sleep <- read_xpt("../Data/Source/SLEEP12.XPT")
```

# NHANES Setup - Merging Data

```{r, echo=FALSE}
df <- lead %>% 
  left_join(., demo, by = "SEQN") %>%
  left_join(., BP, by = "SEQN") %>%
  left_join(., BMI, by = "SEQN") %>%
  left_join(., smoke, by = "SEQN") %>%
  left_join(., ALC, by = "SEQN") %>%
  left_join(., stress, by = "SEQN") %>% 
  left_join(., sleep, by = "SEQN")
```

# NHANES Setup - Defining nhanes and Defining the Outcome

```{r}
nhanes <- df %>%  
  transform(hyper = ifelse(((BPXSY1 >= 130) | (BPXDI1 >= 80)), 1, 0)) #Find a source for the definition of hypertension (n = 8956)
```

Note: Definition of hypertension (according to AHA): https://www.heart.org/en/health-topics/high-blood-pressure/the-facts-about-high-blood-pressure

# NHANES Setup - Data Correction so Never Smokers Don't Automatically Get Dropped

```{r}
nhanes$SMQ040[nhanes$SMQ020 == 2] <- 3
```

# NHANES Setup - Applying Exclusion Criteria and Filtering out NA Values

```{r}
vars <- c("SEQN", "LBXBPB", "RIDAGEYR", "RIAGENDR", "RIDRETH1", "DMDEDUC2", "BMXBMI", "SMQ020", "SMQ040", "ALQ130", "BPXSY1", "BPXDI1", "HSQ480", "SLD010H", "hyper")

nhanes <- nhanes %>% 
  subset(select = all_of(vars)) %>%
  filter(complete.cases(.)) %>%
  filter(RIDAGEYR >= 20)

# nhanes
```

Note: Sleep duration is capped out at 12 hours a day (anything above 12 was recorded as 12)

# NHANES Setup - Filtering out Other Missing Values

```{r}
# Filtering out Lead Limit Below Detection (Limit is 0.25)
nhanes <- nhanes %>% 
  filter(LBXBPB >= 0.25) # (6)
  
# Filtering out Refused/Don't Know Education
nhanes <- nhanes %>% 
  filter(DMDEDUC2 != 7, DMDEDUC2 != 9) # (0)

# Filtering out Refused/Don't Know Smoking History
nhanes <- nhanes %>% 
  filter(SMQ020 != 7, SMQ020 != 9) %>%  
  filter(SMQ040 != 7, SMQ040 != 9) # (1)
  
# Filtering out Refused/Don't Know Alcohol History
nhanes <- nhanes %>% 
  filter(ALQ130 != 777, ALQ130 != 999) # (4)

# Filtering out Refused/Don't Know Stress Levels
nhanes <- nhanes %>% 
  filter(HSQ480 != 77, HSQ480 != 99) # (1)

# Filtering out Refused/Don't Know Sleep Duration
nhanes <- nhanes %>% 
  filter(SLD010H != 77, SLD010H != 99) # (2)

# nhanes (n = 2823)
```

## NHANES Setup - Smoking Data Transformations

```{r}
nhanes <- nhanes %>% 
  mutate(DMDEDUC2 = DMDEDUC2 %>% 
    replace(. == 2, 1) %>%
    replace(. == 4, 5)
  )

nhanes <- nhanes %>% 
  mutate(
    SMOKE = case_when(
      (SMQ020 == 1) & ((SMQ040 == 1) | (SMQ040 == 2)) ~ 1,
      (SMQ020 == 1) & (SMQ040 == 3) ~ 2,
      (SMQ020 == 2) ~ 3
    )
  )
```

# Table 1 Part 1 (NHANES)

```{r}
table1_1 <- data.table('Group' = c(), 'Variable' = c(), 'Nperc' = c())

add_categorical_row <- function(variable, label, value) {
  table1_1 <- table1_1 %>% rbind(data.table('Group' = variable, 
                                        'Variable' = label,
                                        'Nperc' = sum(nhanes[[variable]] %in% value)))
  return(table1_1)
}

table1_1 <- add_categorical_row("RIAGENDR", "Male", 1)
table1_1 <- add_categorical_row("RIAGENDR", "Female", 2)
table1_1 <- add_categorical_row("RIDRETH1", "Mexican American", 1)
table1_1 <- add_categorical_row("RIDRETH1", "Other Hispanic", 2)
table1_1 <- add_categorical_row("RIDRETH1", "Non-Hispanic White", 3)
table1_1 <- add_categorical_row("RIDRETH1", "Non-Hispanic Black", 4)
table1_1 <- add_categorical_row("RIDRETH1", "Other Race - Including Multi-Racial", 5)
table1_1 <- add_categorical_row("DMDEDUC2", "< High School", 1)
table1_1 <- add_categorical_row("DMDEDUC2", "High School", 3)
table1_1 <- add_categorical_row("DMDEDUC2", "> High School", 5)
table1_1 <- add_categorical_row("SMOKE", "Current Smoker", 1)
table1_1 <- add_categorical_row("SMOKE", "Former Smoker", 2)
table1_1 <- add_categorical_row("SMOKE", "Never Smoker", 3)
table1_1 <- add_categorical_row("hyper", "Doesn't Have Hypertension", 0)
table1_1 <- add_categorical_row("hyper", "Has Hypertension", 1)

# sum(data[["hyper"]] == 1)

# data$hyper

table1_1 <- table1_1 %>% group_by(Group) %>% 
  mutate(Nperc = paste0(Nperc, " (", round(Nperc/sum(Nperc), 3)*100, ")")) %>% 
  rename('N (%)' = Nperc) %>% 
  ungroup %>% 
  subset(select = -Group)

## WHICH ROWS TO DELETE???

table1_1
```

# Table 1 Part 2 (NHANES)

```{r}
table1_2 <- data.table('Variable' = c(), 'Mean (std. dev.)' = c())

add_continuous_row <- function(variable, label) {
  table1_2 <- table1_2 %>% rbind(data.table('Variable' = label, 
                                        'Mean (std. dev.)' = paste0(round(mean(nhanes[[variable]]), 2), ' (', round(sd(nhanes[[variable]]), 2), ')')))
  return(table1_2)
}

table1_2 <- add_continuous_row("LBXBPB", "Lead Level (\u03BCg/dL)")
table1_2 <- add_continuous_row("RIDAGEYR", "Age (yrs)")
table1_2 <- add_continuous_row("BMXBMI", "BMI (kg/m\u00B2)")
table1_2 <- add_continuous_row("ALQ130", "Alcoholic drinks per day")
table1_2 <- add_continuous_row("BPXSY1", "Systolic BP (mmHg)")
table1_2 <- add_continuous_row("BPXDI1", "Diastolic BP (mmHg)")
table1_2 <- add_continuous_row("HSQ480", "Days Stressed Per Month")
table1_2 <- add_continuous_row("SLD010H", "Sleep Duration (hrs)")
table1_2
```

## BRFSS

# Overview

Codebooks:
- BFRSS: https://www.cdc.gov/brfss/annual_data/annual_2022.html

List of possible EMM on the RD scale also found in NHANES:
- Stress: MENTHLTH
- Sleep duration (hours of sleep): SLEPTIM1

# BRFSS Setup - Importing Data

```{r}
brfss <- read_xpt("../Data/Derived/BRFSS_derived.xpt")
```

# BRFSS Setup - Applying Exclusion Criteria and Filtering out NA Values

```{r}
brfss <- brfss %>% 
  filter(complete.cases("AGE80", "MENTHLTH", "SLEPTIM1")) %>% # (0)
  filter(AGE80 >= 20) # (6924)
```

Note: Age had been imputed.

# BRFSS Setup - Filtering out Other Missing Values

```{r}
# Filtering out Unknown Mental Health 
brfss <- brfss %>% 
  filter(MENTHLTH != 77, MENTHLTH != 88, MENTHLTH != 99) # (271727)

# Filtering out Unknown Sleep Duration
brfss <- brfss %>% 
  filter(SLEPTIM1 != 77, SLEPTIM1 != 99) # (1680)

# brfss # (n = 164801)
```

# Table 1 Part 3 (BRFSS)

```{r}
table1_3 <- data.table('Variable' = c(), 'Value' = c())

add_continuous_row_3 <- function(variable, label) {
  table1_3 <- table1_3 %>% rbind(data.table('Variable' = label, 
                                        'Value' = paste0(round(mean(brfss[[variable]]), 2), ' (', round(sd(brfss[[variable]]), 2), ')')))
  return(table1_3)
}

add_categorical_row_brfss <- function(variable, label, value) {
  table1_3 <- table1_3 %>% rbind(data.table('Variable' = label,
                                        'Value' = sum(brfss[[variable]] %in% value)/nrow(brfss)))
  return(table1_3)
}

table1_3 <- add_continuous_row_3("AGE80", "Age")
table1_3 <- add_continuous_row_3("MENTHLTH", "Days Stressed per Month")
table1_3 <- add_continuous_row_3("SLEPTIM1", "Sleep Duration (hrs)")
table1_3
```

# Adding More Variables to Table1_3

```{r}
# BMI

old_brfss <- brfss
brfss <- brfss %>% 
  mutate(BMI = BMI/100) %>% 
  filter(!is.na(BMI))
table1_3 <- add_continuous_row_3("BMI", "BMI")
brfss <- old_brfss

# Alcoholic Drinks per Day

old_brfss <- brfss
brfss <- brfss %>% 
  filter(ALCDAY4 != 777, ALCDAY4 != 888, ALCDAY4 != 999, !is.na(ALCDAY4))
table1_3 <- add_continuous_row_3("ALCDAY4", "Alcoholic Drinks per Day")
brfss <- old_brfss

# Sex

old_brfss <- brfss
brfss <- brfss %>% filter(!is.na(SEXVAR))
table1_3 <- add_categorical_row_brfss("SEXVAR", "Male", 1)
table1_3 <- add_categorical_row_brfss("SEXVAR", "Female", 2)
brfss <- old_brfss

# Education

old_brfss <- brfss
brfss <- brfss %>% 
  filter(EDUCA != 9, !is.na(EDUCA))
table1_3 <- add_categorical_row_brfss("EDUCA", "< High School", 1:3)
table1_3 <- add_categorical_row_brfss("EDUCA", "High School", 4)
table1_3 <- add_categorical_row_brfss("EDUCA", "> High School", 5:6)
brfss <- old_brfss

# Smoking Status

old_brfss <- brfss
brfss <- brfss %>% 
  filter(SMOKER != 9, !is.na(SMOKER))
table1_3 <- add_categorical_row_brfss("SMOKER", "Current Smoker", 1:2)
table1_3 <- add_categorical_row_brfss("SMOKER", "Former Smoker", 3)
table1_3 <- add_categorical_row_brfss("SMOKER", "Never Smoker", 4)
brfss <- old_brfss

table1_3
```

## NHANES Analysis

# NHANES Conversion of Some Variables to Factor Type

```{r}
nhanes <- nhanes %>% 
  mutate(RIAGENDR = as.factor(RIAGENDR)) %>% 
  mutate(RIDRETH1 = as.factor(RIDRETH1)) %>% 
  mutate(DMDEDUC2 = as.factor(DMDEDUC2)) %>% 
  mutate(SMOKE = as.factor(SMOKE))
  # select(-c(SEQN, SMQ020, SMQ040, BPXSY1, BPXDI1, ))
```

# Testing for EMM

```{r}
# stress_EMM_model <- glm(hyper ~ LBXBPB + RIDAGEYR + RIAGENDR + RIDRETH1 + DMDEDUC2 + BMXBMI + SMOKE + ALQ130 + LBXBPB*HSQ480, nhanes, family=binomial(link="logit"))

stress_EMM_model <- lm(hyper ~ LBXBPB + RIDAGEYR + RIAGENDR + RIDRETH1 + DMDEDUC2 + BMXBMI + SMOKE + ALQ130 + LBXBPB*HSQ480, nhanes)
summary(stress_EMM_model)

# sleep_EMM_model <- glm(hyper ~ LBXBPB + RIDAGEYR + RIAGENDR + RIDRETH1 + DMDEDUC2 + BMXBMI + SMOKE + ALQ130 + LBXBPB*SLD010H, nhanes, family=binomial(link="logit"))

sleep_EMM_model <- lm(hyper ~ LBXBPB + RIDAGEYR + RIAGENDR + RIDRETH1 + DMDEDUC2 + BMXBMI + SMOKE + ALQ130 + LBXBPB*SLD010H, nhanes)
summary(sleep_EMM_model)
```

P for Interactions:
- Stress = 0.4371
- Sleep Duration = 0.4398

Note that individual terms HSQ480 and SLD010H get added into the regression.

# Testing EMM Distributions

```{r}
ks.test(nhanes$HSQ480, brfss$MENTHLTH)
ks.test(nhanes$SLD010H, brfss$SLEPTIM1)
```

The Kolmogorov-Smirnov test indicates beyond resasonable doubt there are different distributions in days stressed per month but not sleep duration.

## Transportability

# Choice of V

In theory:

C = All confounders listed earlier
Differing variables = {stress}
EMM = {sleep}
V = {} (empty set)

However, for the purposes of illustration, we should still consider sleep as a differing variable and stress as an EMM. So we let V = {stress, sleep}.

# Prepping Study Population Dataframe

```{r}
study <- nhanes %>% 
  rename(x = LBXBPB) %>% 
  rename(c1 = RIDAGEYR) %>% 
  rename(c2 = RIAGENDR) %>% 
  rename(c3 = RIDRETH1) %>% 
  rename(c4 = DMDEDUC2) %>% 
  rename(c5 = BMXBMI) %>% 
  rename(c6 = ALQ130) %>% 
  rename(c7 = SMOKE) %>% 
  rename(v1 = HSQ480) %>% 
  rename(v2 = SLD010H) %>% 
  rename(y = hyper) %>% 
  select(-c(SEQN, SMQ020, SMQ040, BPXSY1, BPXDI1)) %>% 
  mutate(xv1 = x*v1) %>% 
  mutate(xv2 = x*v2)

# study
```

# Prepping Target Population Dataframe

```{r}
target <- brfss %>% 
  rename(v1 = MENTHLTH) %>% 
  rename(v2 = SLEPTIM1) %>% 
  select(-c(AGE80))
            
# target
```

# Simulating Data (Study and Target Pop.)

```{r}
sample_ate_logit <- c()
transported_ate_logit <- c()

N = nrow(target)

set.seed(12345)

for (i in 1:1000) {

study_boot <- copy(study[sample(1:nrow(study),size=nrow(study),replace=TRUE),])  
  
## STUDY POPULATION ATE
  
## Q model, using logistic regression
mod1_l <- glm(data=study_boot, formula="y ~ x + c1 + c2 + c3 + c4 + c5 + c6 + c7 + v1 + v2 + xv1 + xv2", family=binomial(link="logit"))
# summary(mod1_l)
  
## increasing exposure by 1
study1 <- copy(study_boot[,c("x","c1","c2","c3","c4","c5","c6","c7","v1","v2", "xv1", "xv2")])
study1$x <- study1$x + 1
study1$xv1 <- study1$x*study1$v1
study1$xv2 <- study1$x*study1$v2
  
## predict new Y under this intervention 
study1$YdoX_l <- predict(mod1_l,newdata=study1,type="response") ## predict from logistic regression instead (but predict probability)
  
## SATE
sample_ate_logit <- c(sample_ate_logit, mean(study1$YdoX_l-study$y))

## TRANSPORTED ATE

# creating target dataframe

target_temp <- target
target <- study
while (nrow(target) < nrow(brfss)) {
  target <- rbind(target, target)
}
target <- target[1:nrow(brfss), ]

target$v1 <- target_temp$v1
target$v2 <- target_temp$v2

target$xv1 <- target$x*target$v1
target$xv2 <- target$x*target$v2
# target

# Note, I can create arbitrary values of C in the target population - you can set this to anything because the C's cancel out 

## take q-model from study population and predict potential outcome
target$Y_l <- predict(mod1_l,newdata=target,type="response") ## from alternate q-model
  
## now make second target population under intervention, altering X
target1 <- copy(target[,c("x","c1","c2","c3","c4","c5","c6","c7","v1","v2","xv1","xv2")])
target1$x <- target1$x + 1
target1$xv1 <- target1$x*target1$v1
target1$xv2 <- target1$x*target1$v2
  
## again, take q-model from study population and predict potential outcome
target1$YdoX_l <- predict(mod1_l,newdata=target1,type="response") ## from alternate q-model (logit)
  
# transported_ate <- c(transported_ate,mean(target1$YdoX-target$Y))
transported_ate_logit <- c(transported_ate_logit, mean(target1$YdoX_l-target$Y_l))

}
```

# Checking out the Sample ATE and the Transported ATE Point Estimnates and 95% CIs (Table 2)

```{r}
mean(sample_ate_logit)
mean(transported_ate_logit)

quantile(sample_ate_logit,probs=c(0.025,0.975))
quantile(transported_ate_logit,probs=c(0.025,0.975))

# sample_ate_logit
# transported_ate_logit
```

